### 2013-07-24 doc/dissertation/document/chapters/conclusion/conclusion.tex
finished conclusion draft

Throughout this work, three salient issues were observed over all
others. First, the performance of the MCSA implementation developed
for this work was severely lacking in terms of time to solution when
compared to production Krylov methods. In all cases, this difference
was several orders of magnitude. Second, the preconditioning strategy
developed by this work, although often successful in achieving
convergence for difficult problems, creates many obstacles to a
production implementation. These obstacles all primarily stem from the
dense composite operators that for the probabilities and weights for
the Monte Carlo game. Third, perhaps the most restrictive and
unattractive piece of MCSA as a whole is the simple fact that the
spectral radius of the physics operator iteration matrix be less than
one. This was demonstrated to prohibit convergence for cases where the
preconditioning was considered agressive and Krylov solvers exhibited
excellent convergence properties. We will address each one of these
problems in turn in this section and provide suggestions for their
solution.

@@

A second mode of optimization to improve performance regards how
discrete probability distribution functions are sampled in the
implementation. At each transition in the random walk sequence, the
discrete probability distribution function generated by the current
row of the physics operator in which a random walk resides is sampled
to get the next state in the sequence. This operation is also used
when the right hand side of the system is sampled to acquire the
starting state for each random walk in the problem through
Eq~(\ref{eq:adjoint_source_probability}). The current implementation
uses the inverse transform method for sampling where the a random
number is generated and the cumulative distribution function is search
to find the next discrete state. This algorithm has $O(log(N))$ time
complexity for sampling and $O(N)$ time complexity for
construction. The alias sampling method is suggested instead as it
produces $O(1)$ complexity for sampling and $O(N)$ time complexity for
construction. Profiling of the implementation shows that this
improvement could also provide quality reductions in run time.

@@

In general, preconditioning developments in this work enabled much of
the desired research to be performed but also prevented some of it. Of
primary concern is the large memory footprint induced by the explicit
algebraic preconditioning strategy outlined in
\S~\ref{subsubsec:general_mcsa_preconditioning}. By preconditioning
in this way, we were able to leverage all modern algebraic
preconditioning strategies used in current physics practice. The
downside is that in order to generate weights and probabilities with
which to play the Monte Carlo game, me must first invert each of the
preconditioners which in many cases means applying their inverse over
each row in the system, and then we must perform matrix-matrix
multiply operations to build the composite physics operator.

These operations were observed to substantially increase runtimes of
MCSA and consume unacceptable amounts of memory, mostly due to the
fact that inverting the preconditioners create dense
matrices. Although a good fraction of the elements in each row of the
composite operator could be eliminated through the reduced domain
approximation developed by this work, memory costs still prevented
higher fidelity simulations from being performed. In addition, the
reduced domain approximation does no prevent the large run times
incurred for inverting the preconditioners. We therefore propose the
following three developments:

\begin{itemize}
\item \textbf{Research and develop alternative sampling schemes for
  the general left/right preconditioned Monte Carlo method not relying
  on composite matrix generation.} Developments that break away from
  Matrix-Matrix multiplications within the preconditioned solver
  framework improve scalability. This means avoiding the generation of
  the composite operator and instead playing the Monte Carlo game with
  weights and probabilities formed by considering the original linear
  operator and the preconditioners individually. This individual
  consideration may come from an alternative algebraic formulation of
  the general preconditioned MCSA method in
  Eq~(\ref{eq:left_right_mcsa}) or from an alternative random sampling
  strategy for the Monte Carlo game.
\item \textbf{Develop Monte Carlo variance reduction based
  preconditioning strategies that do not rely on an algebraic
  preconditioner.}  Constructing the complete inverse of the
  preconditioner is the core time and scalability constraint when
  modern algebraic preconditioners are used. Building the inverse for
  preconditioners where a sparsity pattern is not enforced typically
  results in a dense matrix that destroys scalability for domain
  decomposed parallel computations. Alleviating both of these issues
  by stochastic preconditioning methods will be studied. Decades of
  research and development in Monte Carlo variance reduction for
  particle transport \cite{booth_1994} may be leveraged to form
  similar strategies for linear systems. In the same way that this
  work has been able to adapt modern particle transport strategies for
  parallelism and apply them to linear systems, the same strategy may
  be applied to variance reduction techniques to improve MCSA
  convergence by modifying the weights and probabilities of the Monte
  Carlo game as an effective means of preconditioning.
\item \textbf{Study physics or PDE-based preconditioners for linear
  systems that produce a reduced-order model of the original system to
  which the Monte Carlo method may be applied.} As the Monte Carlo
  method serves to construct the solution correction vector in the
  MCSA method, the Monte Carlo method may be applied to a
  reduced-order physics or PDE-based system that captures the essence
  of the original system and still allows for acceleration of the
  iterative procedure. By reducing the order of the system, we are
  effectively generating a Monte Carlo problem within MCSA in which
  the linear operator is either more sparse, has a better-conditioned
  eigenvalue spectrum, or both. The reduced domain approximation has
  shown that a purely algebraic strategy for reducing the order of the
  Monte Carlo problem can result in improved overall timing
  performance due to increased sparsity at the cost of iterative
  performance as a result of loss of information. A more well informed
  strategy for order reduction via either an intelligent
  simplification of the PDEs that describe the system or a
  simplification of the physics involved may prove to be more
  effective strategy.
\end{itemize}

@@

\subsection{Monte Carlo Methods of the Second Degree}
\label{subsubsec:2_degree_mc}

\subsection{Stochastic Projection Methods}
\label{subsubsec:stochastic_projection_methods}

@@

>>>>

Throughout this work, three salient issues were observed over all
others. First, the performance of the MCSA implementation developed
for this work was severely lacking in terms of time to solution when
compared to production Krylov methods. In all cases, this difference
was several orders of magnitude. Second, the preconditioning strategy
developed by this work, although often successful in achieving
convergence for difficult problems, creates many obstacles to a
production implementation. These obstacles all primarily stem from the
dense composite operators that are required to form the probabilities
and weights for the Monte Carlo game. Third, perhaps the most
restrictive and unattractive piece of MCSA as a whole is the simple
fact that the spectral radius of the physics operator iteration matrix
be less than one. This was demonstrated to prohibit convergence for
cases where the preconditioning was considered agressive and Krylov
solvers exhibited excellent convergence properties. We will address
each one of these problems in turn in this section and provide
suggestions for their solution.

@@

A second mode of optimization to improve performance regards how
discrete probability distribution functions are sampled in the
implementation. At each transition in the random walk sequence, the
discrete probability distribution function generated by the current
row of the physics operator in which a random walk resides is sampled
to get the next state in the sequence. This operation is also used
when the right hand side of the system is sampled to acquire the
starting state for each random walk in the problem through
Eq~(\ref{eq:adjoint_source_probability}). The current implementation
uses the inverse transform method for sampling where the a random
number is generated and the cumulative distribution function is
queried with a binary seach to find the next discrete state. This
algorithm has $O(log(N))$ time complexity for sampling and $O(N)$ time
complexity for construction. The alias sampling method is suggested
instead as it produces $O(1)$ complexity for sampling and $O(N)$ time
complexity for construction \cite{smith_analysis_2005}. Profiling of
the implementation shows that this improvement could also provide
quality reductions in run time.

@@

In general, preconditioning developments in this work enabled much of
the desired research to be performed but also prevented some of it. Of
primary concern is the large memory footprint induced by the explicit
algebraic preconditioning strategy outlined in
\S~\ref{subsubsec:general_mcsa_preconditioning}. By preconditioning in
this way, we were able to leverage modern algebraic preconditioning
strategies used in current physics applications and enable solutions
for both neutronics and fluid flow simulations. The downside is that
in order to generate weights and probabilities with which to play the
Monte Carlo game, we must first invert each of the preconditioners
which in many cases means applying their inverse over each row in the
system, and then we must perform matrix-matrix multiply operations to
build the composite physics operator.

These operations were observed to substantially increase runtimes of
MCSA and consume unacceptable amounts of memory, mostly due to the
fact that inverting the preconditioners created dense
matrices. Although a good fraction of the elements in each row of the
composite operator could be eliminated through the reduced domain
approximation developed by this work, memory costs still prevented
higher fidelity simulations from being performed. In addition, the
reduced domain approximation does not prevent the large run times
incurred for inverting the preconditioners. We therefore propose the
following three developments in preconditioning for Monte Carlo
methods:

\begin{itemize}
\item \textbf{Research and develop alternative sampling schemes for
  the general left/right preconditioned Monte Carlo method not relying
  on composite matrix generation.} Developments that break away from
  matrix-matrix multiplications within the preconditioned solver
  framework improve scalability. This means avoiding the generation of
  the composite operator and instead playing the Monte Carlo game with
  weights and probabilities formed by considering the original linear
  operator and the preconditioners individually. This individual
  consideration may come from an alternative algebraic formulation of
  the general preconditioned MCSA method in
  Eq~(\ref{eq:left_right_mcsa}) or from an alternative random sampling
  strategy for the Monte Carlo game.
\item \textbf{Develop Monte Carlo variance reduction based
  preconditioning strategies that do not rely on an algebraic
  preconditioner.}  Constructing the complete inverse of the
  preconditioner is the core time and scalability constraint when
  modern algebraic preconditioners are used. Building the inverse for
  preconditioners where a sparsity pattern is not enforced typically
  results in a dense matrix that destroys scalability for domain
  decomposed parallel computations. Alleviating both of these issues
  by stochastic preconditioning methods should be studied. Decades of
  research and development in Monte Carlo variance reduction for
  particle transport \cite{booth_1994} may be leveraged to form
  similar strategies for linear systems. In the same way that this
  work has been able to adapt modern particle transport strategies for
  parallelism and apply them to linear systems, the same strategy may
  possibly be applied to variance reduction techniques to improve MCSA
  convergence by modifying the weights and probabilities of the Monte
  Carlo game as an effective means of preconditioning.
\item \textbf{Study physics or PDE-based preconditioners for linear
  systems that produce a reduced-order model of the original system to
  which the Monte Carlo method may be applied.} As the Monte Carlo
  method serves to construct the solution correction vector in the
  MCSA method, the Monte Carlo method may be applied to a
  reduced-order physics or PDE-based system that captures the essence
  of the original system and still allows for acceleration of the
  iterative procedure. By reducing the order of the system, we are
  effectively generating a Monte Carlo problem within MCSA in which
  the linear operator is either more sparse, has a better-conditioned
  eigenvalue spectrum, or both. The reduced domain approximation has
  shown that a purely algebraic strategy for reducing the order of the
  Monte Carlo problem can result in improved overall timing
  performance due to increased sparsity at the cost of iterative
  performance as a result of loss of information. A more well informed
  strategy for order reduction via either an intelligent
  simplification of the PDEs that describe the system or a
  simplification of the physics involved may prove to be a more
  effective strategy.
\end{itemize}

@@

\subsubsection{Monte Carlo Methods of the Second Degree}
\label{subsubsec:2_degree_mc}

Iterative methods of the $n^{th}$ degree produce solutions at each
iteration of the following form:
\begin{equation}
  \ve{x}^{k+1} = F(\ve{A},\ve{b},\ve{x}^{k},\ve{x}^{k-1},\dots,\ve{x}^{k-n+1})\:,
\label{eq:nth_degree_iter}
\end{equation}
where $\ve{A}$ is the linear operator $\ve{b}$ is the right hand side
as defined by Eq~(\ref{eq:linear_problem}). Richardson's iteration is
an iterative method of the first degree such that $\ve{x}^{k+1} =
F(\ve{A},\ve{b},\ve{x}^{k})$. To move the Monte Carlo method away from
the spectral radius limitation and therefore we seek higher order
methods as outlined by Hallett \cite{hughes_hallett_second-order_1984}
that have garanteed convergence at all eigenvalue spectra. In his
work, Hallett builds a second order method such that $\ve{x}^{k+1} =
F(\ve{A},\ve{b},\ve{x}^{k-1})$. Such a method results from mapping the
original Richardson iteration into a different of eigenvalues in the
complex plane that have a spectral radius less than one, thus
garanteeing convergence. Derived as expansions of the iteration
sequence in Chebyshev polynomials, Hallett arrives at simple
formulation for the second degree method that requires knowledge of
the bounding curve in the complex plane of the eigenvalue spectra for
good convergence properties.

As it turns out, Dimov has already applied such a technique to the
Monte Carlo method but arrived at the same conclusions in a different
way \cite{dimov_parallel_2001,dimov_new_1998}. In his work, Dimov
instead formulated the random walk sequence by instead building the
Neumann series from the \textit{resolvent} of the operator. The
resolvent operator is defined as that which spans all of the complex
plane but has singularities at the points that are eigenvalues of the
original operator. By building the Monte Carlo sequence from this
operator, Dimov arrived at a Monte Carlo method which in fact was
equivalent to a fixed point method of the second degree and utilized
effectively the same spectral mapping procedure as Hallett. Just like
the deterministic case, some knowledge of the eigenvalue spectrum of
the system was required to ensure convergence. These methods should be
researched and developed further in the context of MCSA as a potential
means of always gauranteeing convergence of the algorithm regardless
of the eigenvalue spectrum.

\subsubsection{Stochastic Projection Methods}
\label{subsubsec:stochastic_projection_methods}

As a final means of improvement we consider stochastic projection
methods as developed by Sabelfeld \cite{hutchison_stochastic_2011}. In
his work, Sabelfeld claims his method is 'beyond Markov chains and von
Neumann-Ulam scheme'. To a certain extent this is true, his method is
based on a projection method and should always converge. In his work,
he utilizes a Kaczmarz iteration:
\begin{equation}
  \ve{x}^{k+1} = \ve{x}^k + \lambda_k \frac{b_i - \langle \ve{a}_i
    \ve{x}^k \rangle}{||\ve{a}_i||}\ve{a}_i\:,
  \label{eq:kaczmarz_iteration}
\end{equation}
where $\ve{a}_i$ is the $i^{th}$ row vector of the operator and
$\lambda_k$ is the relaxation parameter.

The stochastic nature of the algorithm comes from effectively
modifying the algorithm such that the value of $i$ is now selected at
random. This is interesting but arguably does not serve our purposes
at well. We have an interest in a fully stochastic process like a
random walk sequence because of its potential benefits in terms of
resiliency and the ability to use superposition to combine results
from replicated problems to enhance performance. In Sabelfeld's work,
the randomness is used to select which deterministic iteration step
will be performed and does not meet either of these
requirements. However, that does not mean there is nothing to take
from this work. If one writes Eq~\ref{eq:kaczmarz_iteration} in terms
of applications of projection operators as that outlined in
\cite{tanabe_projection_1971} we see that this actually takes a
similar form as the Neumann series and could therefore be used in a
random walk sequence. Research studying random walks based on
projection sequences is suggested as a means of moving beyond the
spectral radius limitation.

@@

This work was performed with the goal of investigating new solution
techniques based on Monte Carlo Synthetic Acceleration and how they
perform when applied to real problems in nuclear engineering. These
methods aim to improve over conventional methods when looking towards
the advanced computer architecutures of the near future. Parallelism
at high levels of concurrency, a potential reduction in memory
footprint, and a natural element of resiliency deriving from the
stochastic nature of the algorithm motivated us to research their
strengths and weaknesses and develop solutions to further their
applicability.

Before this work was performed, MCSA had yet to be applied to the
neutron transport problem, asymmetric systems, and nonlinear problems
including fluid flow systems. Techniques were developed by this work
that enabled solutions for all of these problems using MCSA. In many
cases, it was found that the iterative performance of the MCSA-based
methods were superior to that of conventional solution techniques
based on Krylov methods. Furthermore, with improved implementation
performance, future problems may be solved in less time and more
efficiently using MCSA due to this demonstrated improved iterative
performance.

When this work began, the Neumann-Ulam sequence and MCSA in general
had yet to be parallelized in a domain decomposed framework. This
research developed a new parallel algorithm based on Monte Carlo
particle transport that permitted MCSA solutions to be generated on
$O(100,000)$ cores on a leadership class computing facility. The new
algorithm was also demonstrated to achieve better strong and weak
scaling than conventional Krylov methods at high levels of
concurrency. If future work can improve the time to solution using
MCSA through optimization, then this improved scaling demonstrated by
the algorithm can potentially improve the time to solution for the
engineer and more efficiently leverage high performance computing
hardware.

With this work, a new foundation for Monte Carlo Synethic Acceleration
methods has been laid with new theory, algorithms, and numerical
experiments that demonstrate both their potential to make great
contributions to nuclear engineering calculations in the future and
where the research must go in order to achieve this. By advancing
these techniques, we look forward to future work that will enable the
improved design and analysis of nuclear systems.

### 2013-07-24 doc/dissertation/document/chapters/introduction/introduction.tex
finished conclusion draft

In order to leverage Monte Carlo Synthetic Acceleration as a solution
technique for neutron transport problems, its current formulation
requires the physics operator and all preconditioners to be explicitly
formed. Recent developments in the Exnihilo neutronics package at Oak
Ridge National Laboratory have permitted generation of the $SP_N$
system of equations for detailed neutronics models of fission
reactor-based systems\cite{evans_simpli_2013}. By fully forming these
equations and formulating them as a linear algebra problem instead of
using the explicit iterative methods of the past, we now have access
to all of the modern advancements in computational linear algebra
including Krylov solvers for asymmetric systems and algebraic
preconditioning methods.

>>>>

In order to leverage Monte Carlo Synthetic Acceleration as a solution
technique for neutron transport problems, its current formulation
requires the physics operator and all preconditioners to be explicitly
formed. Recent developments in the Exnihilo neutronics package at Oak
Ridge National Laboratory have permitted generation of the $SP_N$
system of equations for detailed neutronics models of fission
reactor-based systems\cite{evans_simplified_2013}. By fully forming these
equations and formulating them as a linear algebra problem instead of
using the explicit iterative methods of the past, we now have access
to all of the modern advancements in computational linear algebra
including Krylov solvers for asymmetric systems and algebraic
preconditioning methods.

### 2013-07-24 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
finished conclusion draft

In this chapter, we briefly derive the simplified $P_N$ ($SP_N$)
equations, closely following the work of Evans
\cite{evans_simpli_2013}\footnote{The $SP_N$ derivations in this
  chapter and the associated appendices are heavily based on those
  presented by Evans in \cite{evans_simpli_2013}}, in order to gain
full understanding of the underlying system and its potential behavior
in a Monte Carlo context. From the $P_N$ equations, we apply a set of
approximations to yield the $SP_N$ equations for fixed source and
criticality problems. Using the fully-formed linear operator for the
transport problem, we explore solutions to the $SP_N$ equations with
Monte Carlo Synthetic Acceleration using a challenging light water
reactor fuel assembly criticality calculation as the driving
problem. Several difficulties arise when applying MCSA to these
problems that were not observed when investigating the simpler
transport systems.

@@

'pseudo-moment' $u_n$. As noted by Evans \cite{evans_simpli_2013},

@@

\subsection{Eigenvalue Form of the $SP_N$ Equations}
\label{subsec:eigenvalue_form}
For criticality problems, the space, angle, and energy discretizations
for the $SP_N$ equations can be applied to the general eigenvalue form
of the transport equation given by
Eq~(\ref{eq:eigenvalue_transport}). We can readily replace the fixed
source in Eq~(\ref{eq:spn_matrix}) with a fission source this time
considering the multigroup form of the $SP_N$ equations derived in
Appendix~\ref{chap:mg_spn_equations}:
\begin{multline}
  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
  \label{eq:multigroup_spn_eigenvalue}
\end{multline}
with the fission matrix $\mathbf{F}$ given by
\cite{evans_simpli_2013}. We again apply the change of variables to
yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
the application of the fission matrix to the moment vectors will be
expanded into a set of block matrices in identical fashion to the
scattering matrices:
\begin{equation}
  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
  \mathbb{F}_{nm} \mathbb{U}_n\ \ \ \ \ \ \ n = 1,2,3,4\:.
  \label{eq:spn_fission_matrix}
\end{equation}
It should be noted here that with respect to a general MCSA scheme,
the introduction of the fission in the system will only affect the
source vector in the linear system. The linear operator, and therefore
overall MCSA performance will be dictated by streaming and scattering
as defined on the left-hand side.

>>>>

In this chapter, we briefly derive the simplified $P_N$ ($SP_N$)
equations, closely following the work of Evans
\cite{evans_simplified_2013}\footnote{The $SP_N$ derivations in this
  chapter and the associated appendices are heavily based on those
  presented by Evans in \cite{evans_simplified_2013}}, in order to gain
full understanding of the underlying system and its potential behavior
in a Monte Carlo context. From the $P_N$ equations, we apply a set of
approximations to yield the $SP_N$ equations for fixed source and
criticality problems. Using the fully-formed linear operator for the
transport problem, we explore solutions to the $SP_N$ equations with
Monte Carlo Synthetic Acceleration using a challenging light water
reactor fuel assembly criticality calculation as the driving
problem. Several difficulties arise when applying MCSA to these
problems that were not observed when investigating the simpler
transport systems.

@@

'pseudo-moment' $u_n$. As noted by Evans \cite{evans_simplified_2013},

@@

\subsection{Eigenvalue Form of the $SP_N$ Equations}
\label{subsec:eigenvalue_form}
For criticality problems, the space, angle, and energy discretizations
for the $SP_N$ equations can be applied to the general eigenvalue form
of the transport equation given by
Eq~(\ref{eq:eigenvalue_transport}). We can readily replace the fixed
source in Eq~(\ref{eq:spn_matrix}) with a fission source this time
considering the multigroup form of the $SP_N$ equations derived in
Appendix~\ref{chap:mg_spn_equations}:
\begin{multline}
  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
  \label{eq:multigroup_spn_eigenvalue}
\end{multline}
with the fission matrix $\mathbf{F}$ given by
\cite{evans_simplified_2013}. We again apply the change of variables to
yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
the application of the fission matrix to the moment vectors will be
expanded into a set of block matrices in identical fashion to the
scattering matrices:
\begin{equation}
  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
  \mathbb{F}_{nm} \mathbb{U}_n\ \ \ \ \ \ \ n = 1,2,3,4\:.
  \label{eq:spn_fission_matrix}
\end{equation}
It should be noted here that with respect to a general MCSA scheme,
the introduction of the fission in the system will only affect the
source vector in the linear system. The linear operator, and therefore
overall MCSA performance will be dictated by streaming and scattering
as defined on the left-hand side.
